**Application of database-independent approach to assess the quality of OTU picking methods**

### 2016-03-25 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Responses to reviewer comments

The recent analysis of Kopylova et al [-@Kopylova2016] repeated many of the benchmarking strategies employed by previous researchers. Many algorithm developers have clustered sequences from simulated communities or sequencing data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016]. This approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome. Furthermore, it is unclear how the methods scale to sequences from the novel organisms we are likely to encounter in deep sequencing surveys. In a second approach, developers have compared the time and memory required to cluster sequences in a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014]. These are valid parameters to assess when judging a clustering method, but indicate little regarding the quality of the OTU assignments. One could randomly assign sequences to a predetermined number of OTUs. This would be efficient. It would also poorly reflect the genetic diversity within the community. In a third approach, developers have compared the number of OTUs generated by various methods for a common dataset [@Edgar2013; @Kopylova2016]. Again, one could randomly cluster sequences into a target number of OTUs, but the clusterings would likely be meaningless. In a fourth approach a metric of OTU stability has been proposed as a way to assess algorithms [@He2015]. Although it is important that the methods generate reproducible OTU assignments when the initial order of the sequences is randomized, this metric ignores the possibility that the variation in assignments may be equally robust or that the assignments by a highly reproducible algorithm may be quite poor. In a final approach, some developers have assessed the quality of clustering based on the method's ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014]. Unfortunately, without the ability to ground truth any method, such comparisons are tenuous. There is a need for an objective metric to assess the quality of OTU assignments.

@@

I replicated the benchmarking approach that I have used previously to assess the ability of an algorithm to correctly group sequences that are similar to each other and split sequences that are dissimilar to each other using the MCC [@Westcott2015; @Schloss2011Assessing]. When I compared the MCC values calculated using the ten *de novo* algorithms with the four datasets, the average neighbor algorithm reliably performed as well or better than the other methods (Figure 1). For the murine dataset, the MCC values for the VSEARCH (AGC: `r mouse["vagc"]` and DGC: `r mouse["vdgc"]`) and USEARCH-based (AGC: `r mouse["agc"]` and DGC: `r mouse["dgc"]`) algorithms, Sumaclust (`r mouse["sumaclust"]`), and average neighbor (`r mouse["an"]`) were similarly high. For each of the other datasets, the MCC value for the average neighbor algorithm was at least 5% higher than the next best method. Swarm does not use a traditional distance-based criteria to cluster sequences into OTUs and instead looks for natural subnetworks in the data. When I used the distance threshold that gave the best MCC value for the Swarm data, the MCC values were generally not as high as they were using the average neighbor algorithm. The one exception was for the soil dataset. Among the reference-based methods, all of the MCC values suffer because when sequences that are at least 97% similar to a reference are pooled, the sequences within an OTU could be as much as 6% different from each other. The effect of this is observed in the MCC values that were calculated for the OTUs assigned by these methods generally being lower than those observed using the *de novo* approaches (Figure 1). It is also important to note that the MCC values for the closed-reference OTUs are inflated because sequences were removed from the analysis if there was not a reference sequence that was more than 97% similar to the sequence. Given the consistent quality of the clusterings formed by the average neighbor algorithm, these results confirm the conclusion from the previous analysis that researchers should use the average neighbor algorithm or calculate MCC values for several methods and use the clustering that gives the best MCC value [@Westcott2015].

>>>>

The recent analysis of Kopylova et al [-@Kopylova2016] repeated many of the benchmarking strategies employed by previous researchers. Many algorithm developers have clustered sequences from simulated communities or sequencing data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016]. Although an OTU definition would ideally match bacterial taxonomy, bacterial taxonomy has proven itself to be fluid and reflect the biases of various research interests. Furthermore, it is unclear how the methods scale to sequences from the novel organisms we are likely to encounter in deep sequencing surveys. In a second approach, developers have compared the time and memory required to cluster sequences in a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014]. These are valid parameters to assess when judging a clustering method, but indicate little regarding the quality of the OTU assignments. For example, reference-based methods are very efficient, but do a poor job of reflecting the genetic diversity within the community when novel sequences are encountered [@Westcott2015]. In a third approach, developers have compared the number of OTUs generated by various methods for a common dataset [@Edgar2013; @Kopylova2016]. Although methods need to guard against excessive splitting of sequences across OTUs, by focusing on minimizing the number of OTUs in a community developers excessively lumping sequences that do not belong together. In a fourth approach a metric of OTU stability has been proposed as a way to assess algorithms [@He2015]. Although it is important that the methods generate reproducible OTU assignments when the initial order of the sequences is randomized, this metric ignores the possibility that the variation in assignments may be equally robust or that the assignments by a highly reproducible algorithm may be quite poor. In a final approach, some developers have assessed the quality of clustering based on the method's ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014]. Unfortunately, without the ability to ground truth any method, such comparisons are tenuous. There is a need for an objective metric to assess the quality of OTU assignments.

@@

I replicated the benchmarking approach that I have used previously to assess the ability of an algorithm to correctly group sequences that are similar to each other and split sequences that are dissimilar to each other using the MCC [@Westcott2015; @Schloss2011Assessing]. When I compared the MCC values calculated using the ten *de novo* algorithms with the four datasets, the average neighbor algorithm reliably performed as well or better than the other methods (Figure 1). For the murine dataset, the MCC values for the VSEARCH (AGC: `r mouse["vagc"]` and DGC: `r mouse["vdgc"]`) and USEARCH-based (AGC: `r mouse["agc"]` and DGC: `r mouse["dgc"]`) algorithms, Sumaclust (`r mouse["sumaclust"]`), and average neighbor (`r mouse["an"]`) were similarly high. For each of the other datasets, the MCC value for the average neighbor algorithm was at least 5% higher than the next best method. Swarm does not use a traditional distance-based criteria to cluster sequences into OTUs and instead looks for natural subnetworks in the data. When I used the distance threshold that gave the best MCC value for the Swarm data, the MCC values were generally not as high as they were using the average neighbor algorithm. The one exception was for the soil dataset. Among the reference-based methods, all of the MCC values suffer because when sequences that are at least 97% similar to a reference are pooled, the sequences within an OTU could be as much as 6% different from each other. The effect of this is observed in the MCC values that were calculated for the OTUs assigned by these methods generally being lower than those observed using the *de novo* approaches (Figure 1). It is also important to note that the MCC values for the closed-reference OTUs are inflated because sequences were removed from the analysis if there was not a reference sequence that was more than 97% similar to the sequence. By choosing to focus on the ability to regenerate taxonomic clusterings, minimizing the number of OTUs, and performance Kopylova et al [-@Kopylova2016] concluded that Swarm and SUMACLUST had the most consistent performance among the *de novo* methods. My objective MCC-based approach found that Sumaclust performed well, but was matched or outperformed by the average neighbor algorithm; using a 3% threshold, Swarm was actually one of the worst methods. Given the consistent quality of the clusterings formed by the average neighbor algorithm, these results confirm the conclusion from the previous analysis that researchers should use the average neighbor algorithm or calculate MCC values for several methods and use the clustering that gives the best MCC value [@Westcott2015].
